---
title: 'Assignment 7: GLMs (Linear Regressions, ANOVA, & t-tests)'
author: "Aislinn McLaughlin"
output:
  pdf_document: default
  word_document: default
geometry: margin=2.54cm
editor_options:
  chunk_output_type: console
---

## OVERVIEW

This exercise accompanies the lessons in Environmental Data Analytics on generalized linear models. 

## Directions
1. Change "Student Name" on line 3 (above) with your name.
2. Work through the steps, **creating code and output** that fulfill each instruction.
3. Be sure to **answer the questions** in this assignment document.
4. When you have completed the assignment, **Knit** the text and code into a single PDF file.
5. After Knitting, submit the completed exercise (PDF file) to the dropbox in Sakai. Add your last name into the file name (e.g., "Fay_A06_GLMs.Rmd") prior to submission.

The completed exercise is due on Tuesday, March 2 at 1:00 pm.

## Set up your session 
1. Set up your session. Check your working directory. Load the tidyverse, agricolae and other needed packages. Import the *raw* NTL-LTER raw data file for chemistry/physics (`NTL-LTER_Lake_ChemistryPhysics_Raw.csv`). Set date columns to date objects.

2. Build a ggplot theme and set it as your default theme.
```{r wrap}

library(knitr)
opts_chunk$set(tidy.opts=list(width.cutoff=50),tidy=TRUE)

```

```{r setup}
#1

library(knitr)
opts_chunk$set(tidy.opts=list(width.cutoff=80),tidy=TRUE)

# setup
getwd()
setwd("~/Documents/GitHub/Environmental_Data_Analytics_2021")

library(tidyverse)
library(agricolae)
library(lubridate)
library(viridis)

# import
lake.chem.phys <-
  read.csv("./Data/Raw/NTL-LTER_Lake_ChemistryPhysics_Raw.csv", stringsAsFactors = TRUE)

# format dates
lake.chem.phys$sampledate <-
  as.Date(lake.chem.phys$sampledate, "%m/%d/%y")

#2

mytheme <-
  theme_gray(base_size = 12) +
  theme(legend.background = element_rect(fill = "gray"), legend.position = "bottom",  plot.title = element_text(face = "bold", size = 14, color = "black", hjust = 0.5), plot.subtitle = element_text(size = 10, color = "gray", hjust = 0.5))

theme_set(mytheme)

```

## Simple regression
Our first research question is: Does mean lake temperature recorded during July change with depth across all lakes?

3. State the null and alternative hypotheses for this question:
> Answer:
H0: Mean lake temperature recorded during July does not change with depth across all lakes (mean lake temperature is the same at all depths during July).
Ha: Mean lake temperature recorded during July does change with depth across all lakes.


4.  Wrangle your NTL-LTER dataset with a pipe function so that the records meet the following criteria: 
 * Only dates in July. 
 * Only the columns: `lakename`, `year4`, `daynum`, `depth`, `temperature_C`
 * Only complete cases (i.e., remove NAs)

5. Visualize the relationship among the two continuous variables with a scatter plot of temperature by depth. Add a smoothed line showing the linear model, and limit temperature values from 0 to 35 째C. Make this plot look pretty and easy to read.

```{r scatterplot}
#4

lake.chem.phys.July <-
  lake.chem.phys %>%
  mutate(month = month(sampledate)) %>%
  filter(month == 7) %>%
  select(lakename, year4, daynum, depth, temperature_C) %>%
  na.omit()

#5

lake.July.plot <-
  ggplot(lake.chem.phys.July, aes(x = temperature_C, y = depth, color = depth)) +
  geom_point() + 
  geom_smooth(method = lm, color = "black") +
  scale_y_reverse() +
  scale_color_viridis(direction = -1) +
  xlim(0, 35) +
  labs(x = "Temperature (째C)", y = "Depth (meters)", title = "Temperature vs. Depth in July")

print(lake.July.plot)

```


6. Interpret the figure. What does it suggest with regards to the response of temperature to depth? Do the distribution of points suggest about anything about the linearity of this trend?

> Answer: As depth decreases, temperature increases. The increase in temperature as it relates to depth accelerates around a depth of 8 m from the surface. The distribution of points suggests this is not a linear relationship. The cluster of points that occur quite far from the line of best fit when depth is high and temperature is low indicate that this is probably more of a logarithmic relationship.


7. Perform a linear regression to test the relationship and display the results

```{r linear.regression}
#7

temperature.regression <-
  lm(data = lake.chem.phys.July, temperature_C ~ depth)

summary(temperature.regression)

```


8. Interpret your model results in words. Include how much of the variability in temperature is explained by changes in depth, the degrees of freedom on which this finding is based, and the statistical significance of the result. Also mention how much temperature is predicted to change for every 1m change in depth. 

> Answer: Changes in depth explain 73.87% of the variability in temperature (simple linear regression, df = 9726, p-value = 2.2e-16). Because our p-value is below 0.05, this regression is meaningful. For every 1m increase in depth, temperature will decrease by ~ 1.95째C.


---

## Multiple regression
Let's tackle a similar question from a different approach. Here, we want to explore what might the best set of predictors for lake temperature in July across the monitoring period at the North Temperate Lakes LTER. 


9. Run an AIC to determine what set of explanatory variables (year4, daynum, depth) is best suited to predict temperature.

10. Run a multiple regression on the recommended set of variables. 

```{r temperature.model}
#9

temp.AIC <-
  lm(data = lake.chem.phys.July, temperature_C ~ year4 + daynum + depth)

step(temp.AIC)

#10

temp.model <- lm(data = lake.chem.phys.July, temperature_C ~ year4 + daynum + depth)
summary(temp.model)

```

11. What is the final set of explanatory variables that the AIC method suggests we use to predict temperature in our multiple regression? How much of the observed variance does this model explain? Is this an improvement over the model using only depth as the explanatory variable?

> Answer: The AIC method recommends we use year, day number and depth to predict temperature in our multiple regression. This new model explains 74.11% of variance in temperature. This is an improvement over our model using depth as the sole explanatory variable because in that model depth only explained 73.87% of the variance in temperature.



---
## Analysis of Variance

12. Now we want to see whether the different lakes have, on average, different temperatures in the month of July. Run an ANOVA test to complete this analysis. (No need to test assumptions of normality or similar variances.) Create two sets of models: one expressed as an ANOVA models and another expressed as a linear model (as done in our lessons).

```{r anova.model}
#12

temp.lakes.July.ANOVA <-
  aov(data = lake.chem.phys.July, temperature_C ~ lakename)
summary(temp.lakes.July.ANOVA)

temp.lakes.July.ANOVA2 <-
  lm(data = lake.chem.phys.July, temperature_C ~lakename)
summary(temp.lakes.July.ANOVA2)

```

13. Is there a significant difference in mean temperature among the lakes? Report your findings. 

> Answer: Our null hypothesis is that the mean temperature is the same across all lakes. Both the ANOVA and linear models give us p-values < 0.05 so we can reject the null hypothesis. There is a statistically significant difference in the means across all lakes.


14. Create a graph that depicts temperature by depth, with a separate color for each lake. Add a geom_smooth (method = "lm", se = FALSE) for each lake. Make your points 50 % transparent. Adjust your y axis limits to go from 0 to 35 degrees. Clean up your graph to make it pretty. 

```{r scatterplot.2}
#14.

temp.depth.lakes.plot <-
  ggplot(data = lake.chem.phys.July, aes(x = depth, y = temperature_C, color = lakename)) +
  geom_point(alpha = 0.5) +
  geom_smooth(method = "lm", se = FALSE) +
  ylim(0, 35) +
  labs(x = "Depth (meters)", y = "Temperature (째C)", title = "Temperature vs Depth", subtitle = "North Temperate Lakes District", color = "Lake Name") +
  guides(color = guide_legend(nrow = 3))

temp.depth.lakes.plot


```

15. Use the Tukey's HSD test to determine which lakes have different means.

```{r tukey.test}
#15

TukeyHSD(temp.lakes.July.ANOVA)
temp.lakes.totals <- 
  HSD.test(temp.lakes.July.ANOVA, "lakename", group = TRUE)
temp.lakes.totals


shapiro.test(lake.chem.phys.July$temperature_C[lake.chem.phys.July$lakename == "Peter Lake"])
shapiro.test(lake.chem.phys.July$temperature_C[lake.chem.phys.July$lakename == "Paul Lake"])

lake.chem.phys.pp <-
  lake.chem.phys.July %>%
  filter(lakename == "Peter Lake" | lakename == "Paul Lake")

var.test(lake.chem.phys.pp$temperature_C ~ lake.chem.phys.pp$lakename)

```

16.From the findings above, which lakes have the same mean temperature, statistically speaking, as Peter Lake? Does any lake have a mean temperature that is statistically distinct from all the other lakes?

>Answer: Paul Lake and Ward Lake do not have significantly different mean temperatures from Peter Lake. No lakes have a mean temperature that is statistically distinct from all the other lakes.
 

17. If we were just looking at Peter Lake and Paul Lake. What's another test we might explore to see whether they have distinct mean temperatures? 

>Answer: We could also use a two-sample t-test to compare the means of Peter Lake and Paul Lake. 


